### 2021-02-16 rad-icfp-2021/abstract.tex
some very rough starting draft

\begin{abstract}
Our work aims at bridging the gap between high-level functional reverse-mode differentiation and efficient imperative implementations.
We define a new source-to-source transformation for reverse-mode differentiation on a standard higher-order purely functional language. 
Using a logical relations argument, we prove our transformation correct using diffeological spaces as a denotational semantics for our language. 
We then show that combining our macro with some classical optimisations lead to low-level SSA-like code, which is as efficient as low-level imperative implementations. 
We prove some theoretical complexity guarantees and provide experimental results which showcase the efficiency of our method. 
\end{abstract}

>>>>

\begin{abstract}
Recent work on automatic differentiation (AD) either focus on expressiveness and correctness, or focus on efficiency. 
Reverse mode implementations are notoriously hard to optimize as they add unpurity to the transformed program, and harder to prove correct or efficient.
We start from a high-level picture on AD where we can prove correctness and keep purity, and show how to turn it into efficient lower level code. 
On a small functional language, we show how the optimized code after AD looks like static single assignment (SSA) code. 
On a more expressive language with higher-order primitives, we introduce an intermediate representation, which we call unary form (UNF), 
which enables us to perform efficient reverse-mode on these constructs while keeping purity and efficiency.
We give a denotational semantics to our language, we show correctness of our reverse-mode transformation, and we prove complexity guarantees. 
\end{abstract}

### 2021-02-16 rad-icfp-2021/background.tex
some very rough starting draft

\section{Pure $\lambda$-calculus for differentiable programming} % (fold)
\label{sec:pure_lambda_calculus_for_differentiable_programming}

\subsection{Source language} % (fold)
\label{sub:source_language}

\subsection{Target Language} % (fold)
\label{sub:target_language}

% subsection target_language (end)

>>>>

\section{Rudiments of AD}

\subsection{Rudiments of differentiation and dual numbers.}
Recall that the derivative of a function $f:\RR\to \RR$, if it exists, is a function
$\nabla f:\RR\to \RR$ such that $\nabla f(x_0)=\frac {\dif f(x)}{\dif x}(x_0)$ is the gradient of $f$ at $x_0$.

To find $\nabla f$ in a compositional way, two generalizations are reasonable:
\begin{itemize}
\item We need both $f$ and $\nabla f$ when calculating $\nabla (f;g)$
of a composition $f;g$, using the chain rule, so we are really interested in the pair $(f,\nabla f):\RR\to \RR\times \RR$;
\item In building $f$ we will need to consider functions of multiple arguments, such as $+:\RR^2\to \RR$, and these functions should propagate derivatives.
\end{itemize}
Thus we are more generally interested in transforming a function $g:\RR^n\to \RR$ into a function
$h:(\RR\times \RR)^n\to \RR\times \RR$ in such a way that for any
$f_1\dots f_n:\RR\to\RR$, 
\begin{equation}
  \label{eqn:dualnumber}
  (f_1,\nabla f_1,\dots, f_n,\nabla f_n);h
  =
  ((f_1,\dots, f_n);g,\nabla ((f_1, \dots, f_n);g))\text.
\end{equation}

An intuition for $h$ is often given in terms of dual numbers.
The transformed function operates on pairs of numbers, $(x,x')$, and it is common
to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
But while this is a helpful intuition, the formalization of infinitesimals can be intricate, 
and the development in this paper is focussed on the elementary formulation in~\eqref{eqn:dualnumber}.

The reader may also notice that $h$ encodes all the partial derivatives of
$g$. For example, 
if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
$h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
and similarly 
$h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
And conversely, if $g$ is differentiable in each argument, then
a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
combinations of partial derivatives:
\[\textstyle h(x_1,x_1',x_2,x_2')=(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))\text.\]

In summary, the idea of differentiation with dual numbers is 
to transform a differentiable function
$g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives. We packaged this up in~\eqref{eqn:dualnumber} as a sort-of invariant which is useful for building derivatives of compound functions $\RR\to\RR$ in a compositional way.
The idea of automatic differentiation is to perform this transformation at the source code level.

The two main ways in which it is performed in practice is by operator overloading \cite{} or by source code transformation. We focus our work on the latter.

\subsection{Reverse-mode AD}

AD comes in two main flavors: forward mode and reverse mode. 
Forward mode computes a directional derivative of the Jacobian while reverse-mode computes a directional derivative of the transpose Jaobian. 
This has a significant impact in practice as this means reverse mode can compute the whole gradient of a function $f:\RR^n\to \RR$ in one pass, whereas forward mode needs $n$ passes. 
Assuming that both have about the same complexity, this can have a huge impact in practice. 
Such functions are expremely commonly seen as loss functions in the context of machine learning for instance.

Reverse-mode computes the derivatives backward, using the symmetry of the chain rule.
This means the compute flow is reverted after the first pass to compute the value of the function and its intermediate values. 
These values appear in the chain rule so should be stored or recomputed.
These things make reverse-mode harder to implement efficiently, and especially harder to analyse and optimise. 
Traditional implementions use references.
For instance, let 
 \begin{align*}
     \trm\defeq &\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
 \end{align*}
 The imperative version of the reverse differentiation of $\trm$ is
	\begin{align*}
		&w_1=\var_1 * \var_2 \\
		&w_2=w_1 * \var_1 \\
        &\var_1':=0~;~\var_2':=0 \\
        &w_1':=0 \\
		&w_2':=1 \\
		&w_1'+= \var_1 * w_2'; \var_1'+= w_1*w_2' \\
		&\var_1'+= \var_2 * w_1'; \var_2'+=\var_1 * w_1'\\
		&\text{return }(\var_1',\var_2')  
	\end{align*}
The tangent part for every variable from the context $\var_i'$ and every intermediate variable $w_i'$ is initialized at $0$, 
except for the return variable $w_2$ for which  $w_2':=1$.

As noted in \cite{pearlmutter2008reverse}, in a functional language, one can have a purely functional reverse-mode using continuations. 
This representation is notoriously inefficient and all the implementions of reverse mode in a functional setting have followed the lines of the imperative version above \cite{pearlmutter2008reverse, wang2018demystifying, baydin2016diffsharp}.

In the next section, we show on a very simple language how we can use standard optimisations to turn this inefficient representation into a very efficient one which is still purely functional.
The remaining of the paper will generalise this idea and explore some consequences of this new approach.

### 2021-02-16 rad-icfp-2021/intro.tex
some very rough starting draft

\MH{might get some inspiration from older notes}

@@

>>>>

@@

\[
\begin{tikzcd}
	& \TSSyn \ar[r,"{\Dsynrev[\rho]{-}}"] 
	& \TTSynHO_1 \ar[d,"\UNF{-}^{-1}"] &  \\
	& \SSSyn \ar[u,"\UNF{-}"]  \ar[r,"{\Dsynrev[\rho]{-}^*}"'] 
	& \TTSynHO_2 \ar[d,"Jacob"] \\
	& \SSSyn \ar[u,"ANF"] \ar[r,"{\Dsynrev[\rho]{-}^{**}}"'] 
	&  \TTSynHO_3 \ar[dl,"\partial_2"] \\
	\SSSynHO \ar[ur,"\partial_1"]  \ar[rrr,bend right=20,color=red,"Example\,1"]
	& \TTSyn \ar[r,"OPT"'] 
	& \TTSyn \ar[r,"\partial_3"']
	& \TTSyn 
\end{tikzcd}
\]

\begin{itemize}
	\item $\SSSynHO$ is a higher-order language and the language the user uses to write a prgram.
	\item $\partial_1$ denotes partial evaluation of all higher-order variables. We end up with a program whose only higher-order bits are primitive operators.
	\item $\SSSyn$ is a first-order language with a few higher-order primitives like map,fold,etc.
	\item $ANF$ is for (weak) A-normal form. Every operator $\op$ is only given variables as arguments. We do this by introducting additional let bindings. It is only used for the $Jacob$ optimisation.
	\item $\UNF{-}$ is the unary form transformation presented in the next section. Its goal is to deal with the fanout problem before applying the reverse-mode macro.
	\item $\TSSyn$ is the target language for $\UNF{-}$ and also the source for the reverse-mode macro. It has a bit of a special structure.
	\item $\Dsynrev[\rho]{-}$ is a reverse-mode macro. It is efficient on unary operators but does not deal well with fanout. Fortunately, it is only given unary operators at this point.
	\item $\TTSynHO_1$ is the target language of the reverse-mode macro. It is a higher-order language as it needs a continuation for the reverse derivatives.
	\item $\TTSynHO_2,\TTSynHO_3$ are small variations of $\TTSynHO_1$.
	\item $\UNF{-}^{-1}$ is an optimisation which removes the overhead caused by $\UNF{-}$. It reverts the primal part of the derivative to the original term.
	\item $\Dsynrev[\rho]{-}^*$ is an efficient macro for reverse-mode which deals with the fanout problem.
	\item $Jacob$ stands for an efficient representation of the Jacobian. It needs a weak form of ANF to compute the Jacobian with a single vector and does not need to use a matrix-vector multiplication. In particular, it does not need to store the matrix.
	\item $\Dsynrev[\rho]{-}^{**}$ is the efficient macro for reverse mode that would really be implemented. It will be quicker to run than the composition of the transformations. 
	\item $\partial_2$ stands for partial evaluation. We evaluate (CBN) the higher-order variables binding the lambdas we introduced and inlined with the reverse-mode macro
	\item $\TTSyn$ is the target language of the previous partial evaluation and the final language we use.
	\item $OPT$ is an optimisation. First, we remplace the previously inlined lambdas by let bindings, and then apply forward substitution.
	\item $\partial_3$  is optionnal. We partially evaluate more and do algebraic simplifications, like $0+\trm\to\trm, 1*\trm\to \trm$.
	\item $Example\,1$ is detailed below.
\end{itemize}

\MH{I could add a different name for the target of ANF, of Jacob, and of OPT.}

\begin{example}
	Let 
\begin{align*}
	\var_1,\var_2,\var_3:\RR\vdash \trm:=\big(\lambda f:\RR\to\RR. (f(\var_1))*\var_1\big)(\lambda \var[2]. \var[2]* \var_2)
\end{align*}
After $\partial_1$ we obtain 
\begin{align*}
	(\letin{\var[2]}{\var_1}{\var[2]*\var_2})*\var_1
\end{align*}
After $ANF$ we obtain
\begin{align*}
	&\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
\end{align*}
After $\Dsynrev[\rho]{-}^*$ we obtain
	\begin{align*}
		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Y\big(J^1(\var_1,\var_2,\var_3)^T(\var[2]_1,...,\var[2]_4)\big)} }{
		\\
		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'\big(J^2(\var_1,\var_2,\var_3,w_1)^T(\var[2]_1,...,\var[2]_5)\big)}}{\\ 
		& w_2'}}
	\end{align*}
where $Y$ is a continuation variable of type $\RR^3\to\RR^3$, $J^1,J^2$ are Jacobian matrices for $*$ operators which also return their context, $\inllambda$ is just a normal $\lambda$ but inlined for future optimisation.

After applying $Jacob$, we obtain
	\begin{align*}
		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Y(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
		\\
		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
		& w_2'}}
	\end{align*}
To actually compute the reverse-derivative of the full term, we substitute the identity to $Y$ and apply the whole term to $(0,0,0,0,1)$, the functional equivalent to initialing the sensitivities in the imperative reverse-mode algorithm. We obtain

\begin{align*}
		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
		\\
		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
		& w_2'(0,0,0,0,1)}}
	\end{align*}

After $\partial_2$, we get
	\begin{align*}
		&\letin{w_1}{\var_1 * \var_2}{\\
		&\letin{w_2}{w_1 * \var_1}{\\
		&\bigg[\inllambda \var[2]_1...\var[2]_5.\Big(\inllambda \var[2]_1'...\var[2]_4'.(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)\Big)(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)\bigg](0,0,0,0,1)
		}
		}
	\end{align*}

	Now let's apply the first part of $OPT$:
	\begin{align*}
		&\letin{w_1}{\var_1 * \var_2}{\\
		&\letin{w_2}{w_1 * \var_1}{\\
		&\letin{(\var[2]_1\ldots\var[2]_5)}{(0,0,0,0,1)}{\\
		& \letin{(\var[2]_1'\ldots\var[2]_4')}{(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}{\\
		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2'+\var_1*\var[2]_4',\var[2]_3')
		}}}}
	\end{align*}
Let's unsugar the lets:
	\begin{align*}
		&\letin{w_1}{\var_1 * \var_2}{\\
		&\letin{w_2}{w_1 * \var_1}{\\
		&\letin{\var[2]_1}{0}{\\
		&\letin{\var[2]_2}{0}{\\
		&\letin{\var[2]_3}{0}{\\
		&\letin{\var[2]_4}{0}{\\
		&\letin{\var[2]_5}{1}{\\
		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
		&\letin{ {\color{red} \var[2]_2'} }{  {\color{red} \var[2]_2} }{\\
		&\letin{ {\color{red} \var[2]_3'} }{ {\color{red} \var[2]_3}  }{\\
		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2'+\var_1*\var[2]_4',\var[2]_3')
		}}}}}}}}}}}
	\end{align*}
	We apply the second part of $OPT$ to remove the variables in red:
	\begin{align*}
		&\letin{w_1}{\var_1 * \var_2}{\\
		&\letin{w_2}{w_1 * \var_1}{\\
		&\letin{\var[2]_1}{0}{\\
		&\letin{\var[2]_2}{0}{\\
		&\letin{\var[2]_3}{0}{\\
		&\letin{\var[2]_4}{0}{\\
		&\letin{\var[2]_5}{1}{\\
		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2+\var_1*\var[2]_4',\var[2]_3)
		}}}}}}}}}
	\end{align*}
	We apply some algebraic simplifications in $\partial_3$
	\begin{align*}
		&\letin{w_1}{\var_1 * \var_2}{\\
		&\letin{w_2}{w_1 * \var_1}{\\
		& \letin{\var[2]_1'}{w_1}{\\
		&\letin{\var[2]_4'}{\var_1}{\\
		&(\var[2]_1'+\var_2*\var[2]_4',\var_1*\var[2]_4',0)
		}}}}
	\end{align*}
	We can optimise some more with $\partial_3$, and we finally obtain
	\begin{align*}
		&\letin{w_1}{\var_1 * \var_2}{\\
		&\letin{w_2}{w_1 * \var_1}{\\
		&(w_1+\var_2*\var_1,\var_1*\var_1,0)
		}}
	\end{align*}
\end{example}
